preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-08 Manuscript/manuscript.Rmd
Abstract and Intro polishing

**Background** High-throughput chromosome conformation capture technology (Hi-C) has revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. TAD and loop boundaries have been characterized by the presence of CTCF and members of cohesin complex binding sites, typically mapped at higher precision. Several methods have been developed to leverage genome annotation  data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary. Furthermore, the number of regions containing boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

**Methods** To improve the detection of domain boundaries, we developed preciseTAD, a random forest-based framework that leverages high-resolution genome annotation data coupled with density-based clustering and scalable partitioning techniques. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to investigate class imbalance. We benchmarked our method against Arrowhead TAD caller and a novel chromatin loop prediction algorithm, Peakachu, on multiple cell lines.

**Results** Using different feature engineering and resampling strategies, we found spatial relationships between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for nearby CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using genome annotation data. Our method is implemented as a preciseTAD R package (Bioconductor URL). The precise identification of TAD boundaries will improve our understanding of how epigenomics shapes the 3D structure of the genome.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), less so for chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of bases [@mendoza2009genome]. Therefore, leveraging such elements in a supervised machine learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at much higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning (ML) framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution genome annotation data in the form of chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel pipeline for building optimal boundary prediction classifier composed of a resolution-specific binning process and feature engineering procedure. We found that spatial associations between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level, *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the 3D architecture of the human genome.

@@

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD caller. Given the wide variety of TAD callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" TAD boundaries. <!--Ideally we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets with boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking.-->Ultra-deep Hi-C sequencing [@bonev2017multiscale] and newer technologies for precise mapping of chromatin interactions (e.g., Micro-C [@Krietenstein:2020aa]), coupled with more precise technologies for genomic annotation profiling (e.g., CUT&RUN for precise mapping of transcription factor binding sites) will help to refine the location and the genomic signatures of "ground truth" TAD boundaries. In the current work, given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

Another limitation of *preciseTAD* is that it does not distinguish boundary types. The hierarchical nature of TAD boundaries [@weinreb2016identification; @fraser2015hierarchical; @Cresswell:2020ab] is not considered by *preciseTAD* due to the lack of gold standard of TAD hierarchy. Recent research distinguishes CTCF-associated boundaries, CTCF-negative YY1-enriched boundaries, CTCF- and YY1- depleted promoter boundaries, and a fourth class of weak boundaries largely depleted of all three features [@Krietenstein:2020aa]. Furthermore, actively transcribed regions can serve as TAD boundaries themselves, independently of CTCF binding [@Harrold:2020aa]. This may lead to some TAD boundaries being undetected by *preciseTAD* despite being detected by TAD callers. Our future work will include defining separate models trained on different TAD boundary types, and predicting precise locations for each of them.

>>>>

**Background** High-throughput chromosome conformation capture technology (Hi-C) has revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, known to play critical roles in cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. TAD and loop boundaries have been characterized by the presence of CTCF and members of cohesin complex binding sites, typically mapped at higher precision. Several methods have been developed to leverage genome annotation  data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary. Furthermore, the number of regions containing boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

**Methods** To improve the detection of domain boundaries, we developed preciseTAD, a random forest-based framework that leverages high-resolution genome annotation data coupled with density-based clustering and scalable partitioning techniques. We investigated several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to address class imbalance. We benchmarked our method against Arrowhead domain caller and a novel chromatin loop prediction algorithm, Peakachu, on multiple cell lines.

**Results** Using different feature engineering and resampling strategies, we found spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using genome annotation data. The precise identification of TAD boundaries will improve our understanding of how genomic regulators shapes the 3D structure of the genome.

**Availability.** Our method is implemented as an R package, *preciseTAD*, available on https://github.com/dozmorovlab/preciseTAD and Bioconductor (submitted).

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as domain callers, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [??? Is this the right statement and reference? as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison] ???]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be present at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreths bases [@mendoza2009genome], well below tens of kilobases resolution of Hi-C data (750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of TAD (or loop) boundaries.

Here we present *preciseTAD*, an R package implementing optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines. This improved precision in TAD and loop boundary location can provide insight into the role of genomic regulators on the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined domain caller. Given the wide variety of domain callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" TAD boundaries. <!--Ideally we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets with boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking.-->Ultra-deep Hi-C sequencing [@bonev2017multiscale] and newer technologies for precise mapping of chromatin interactions (e.g., Micro-C [@Krietenstein:2020aa]), coupled with more precise technologies for genomic annotation profiling (e.g., CUT&RUN for precise mapping of transcription factor binding sites) will help to refine the location and the genomic signatures of "ground truth" TAD boundaries. In the current work, given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

Another limitation of *preciseTAD* is that it does not distinguish boundary types. The hierarchical nature of TAD boundaries [@weinreb2016identification; @fraser2015hierarchical; @Cresswell:2020ab] is not considered by *preciseTAD* due to the lack of gold standard of TAD hierarchy. Recent research distinguishes CTCF-associated boundaries, CTCF-negative YY1-enriched boundaries, CTCF- and YY1- depleted promoter boundaries, and a fourth class of weak boundaries largely depleted of all three features [@Krietenstein:2020aa]. Furthermore, actively transcribed regions can serve as TAD boundaries themselves, independently of CTCF binding [@Harrold:2020aa]. This may lead to some TAD boundaries being undetected by *preciseTAD* despite being detected by domain callers. Our future work will include defining separate models trained on different TAD boundary types, and predicting precise locations for each of them.
